High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-05-16 notebooks/mnist.qmd
improved tables

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=Λ[1]),
    "REVISE" => REVISEGenerator(λ=Λ[1]),
    "Schut" => GreedyGenerator(),
    "ECCCo" => eccco_generator,
)
```

>>>>

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=Λ[1]),
    "REVISE" => REVISEGenerator(λ=Λ[1]),
    "Schut" => greedy_generator,
    "ECCCo" => eccco_generator,
)
```

### 2023-05-16 notebooks/tables.Rmd
improved tables

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles", "Linearly Separable")
dt[,source:=ifelse(dataname %in% synth, "synthetic", "real-world")]
```

```{r}
tab <- dt[
    ,
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable, source)
]

@@

# Add conditional formatting:
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)

@@

@@

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
kbl(
  tab_i, caption = caption, 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
kbl(
  tab_i, caption = caption, 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

>>>>

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_benchmark.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles", "Linearly Separable")
dt[,source:=ifelse(dataname %in% synth, "synthetic", "real-world")]
reported_data <- c(
  "MNIST",
  "GMSC",
  "Linearly Separable",
  "Moons",
  "Circles"
)
dt <- dt[dataname %in% reported_data]
dt[,non_valid:=variable=="validity" & value==0.0,.(sample,dataname,generator,model,target,factual,source)]
dt[,non_valid:=any(non_valid==TRUE),.(sample,dataname,generator,model,target,factual,source)]
dt_valid <- dt[non_valid==FALSE]
```

```{r}
tab <- dt[
    ,
    .(
      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable, source)
]

@@

# Add conditional formatting:
tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]

@@

```{r}
tab_valid <- dt_valid[
    ,
    .(
      value=sprintf("%1.2f ± %1.2f", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable, source)
]

tab_valid$top_val = F
tab_valid$one_std_wachter = F
tab_valid$two_std_wachter = F

# Measures to be minimized:
min_measures <- c(
  "distance",
  "distance_from_energy", 
  "distance_from_targets",
  "set_size_penalty"
)
tab_valid[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
tab_valid[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab_valid[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab_valid[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]

# Measures to be maximized:
max_measures <- c(
  "validity",
  "redundancy"
)
tab_valid[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
tab_valid[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab_valid[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
tab_valid[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]

# Add conditional formatting:
tab_valid$value <- cell_spec(tab_valid$value, "latex", bold=tab_valid$top_val)
tab_valid[one_std_wachter==T,value:=paste0(value,"*")]
tab_valid[one_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]
tab_valid[two_std_wachter==T,value:=paste0(value,"*")]
tab_valid[two_std_wachter==F,value:=paste0(value,"\\hphantom{*}")]

# Remove redundant columns:
tab_valid[,val:=NULL]
tab_valid[,std:=NULL]
tab_valid[,top_val:=NULL]
tab_valid[,two_std_wachter:=NULL]
tab_valid[,one_std_wachter:=NULL]
```

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages plus/minus one standard deviation. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages plus/minus one standard deviation. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

@@

## Full table (valid only)

```{r}
tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
kbl(
  tab_full, caption = "All results for all datasets including only valid counterfactuals. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = F, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all_valid.tex")
```
